Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-05-09 manuscript/manuscript.tex
Changes in manuscript

On the variable density tesseroid problem we found two main error 
sources in the GLQ that should be addressed: (a) the ratio between 
the tesseroid size and its distance to the computation point, and (b) 
the variation of the density function.

@@

Because the algorithm will be iteratively applied for every 
subdividing tesseroid, we'd like to note that this normalised density 
function will be the same for every subdividing tesseroid.

In case that the density function is constant, both the maximum and 
minimum densities are equal and the normalised function could not 
be defined, so the density-based discretization algorithm won't be 
applied.

@@

\noindent where the delta ratio $\delta$ is a predefined constant value 
and $L_r^\text{sm}/L_r$ is a tesseroid size ratio, where 
$L_r^\text{sm}$ and $L_r$ are the radial dimensions of the tesseroid 
that is currently analysed and of the \emph{original} one, 
respectively.
On the first iteration, the former tesseroid corresponds to the 
\emph{original} one, so $L_r^\text{sm}/L_r$ is equal to one, but for 
future iterations this radial dimension will be computed as the 
difference of the top and bottom radii of the subdividing tesseroid.
The tesseroid size ratio is intended to reduce the number of probable 
unneeded discretizations and focus on discretizations that will 
significantly reduce the numerical error.
A big subdividing tesseroid with small maximum absolute difference would 
introduce more numerical error than a small subdividing tesseroid with 
higher maximum absolute difference.

@@

On the homogeneous density tesseroids case, \citet{Uieda2016} compared 
their numerical model with the analytical solution of a spherical 
shell in order to obtain default values for the distance-size ratio 
$D$.
We could follow this idea, but for our needs the spherical shell must 
have the same density function as the numerical model.
In contrast with the homogeneous density case, the analytical solutions 
for a variable density spherical shell must be obtained.

@@

The equation \ref{eq:shell-pot} allows to easily obtain the exact 
gravity potential generated by a spherical shell with a variable 
density in depth on any outside point located on the $z$ axe.
Due to the rotational symmetry along any axe that passes through the 
centre of the shell, the expression in equation \ref{eq:shell-pot} 
reproduces the gravity potential field on any outside point at distance 
$r$ from the its centre.

@@

Firstly we need to solve the integral on equation \ref{eq:shell-pot} in 
order to obtain the analytical solution in this case.
Thus, the gravitational potential generated by a spherical shell with this 
density on any external point at distance $r$ from its centre can be 
obtained from the following expression:

@@

\noindent The first term on equation \ref{eq:shell-pot-linear} 
reproduces the potential generated by a spherical shell with variable 
density $\rho(r') = ar'$, while the second one constitutes the 
potential generated by a spherical shell with homogeneous density $\rho 
= b$ \citep{Mikuska2006,Grombein2013}.
So the potential can be read as the one generated by a combination of 
both shells.

Secondly we can analyse how the numerical algorithm works for any 
tesseroid with linear density.
When applying the density-based discretization algorithm on any 
tesseroid with this kind of density, the absolute difference defined in 
equation \ref{eq:density-abs-diff} will always be zero, so the 
discretization inequation \ref{eq:delta-density} will never be 
satisfied.
Therefore, independently of the linear function proposed, the 
density-based algorithm will never be applied, leaving the modified 
adaptive discretization algorithm as the only mechanism to control the 
accuracy of the numerical approximation.
That's why we only need to determine the minimum value of $D$ needed in 
order to guarantee an acceptable accuracy while ignoring the value of 
$\delta$.

@@

We will then compute these gravitational fields on every point of the four 
grids for different values of $D$, ranging from 0.5 to 10 with a step 
of 0.5.
Then we will calculate the maximum difference between these results and the 
value that the analytical solution assumes at the same height of the 
computation grid.
Because the gravity fields generated by a tesseroid are approximated by 
the effect of point masses, the numerical results may vary between 
computation points at same height but at different latitude or 
longitude.
That's why we will choose for comparison the maximum difference between the 
results on every point of the grid and the analytical solution.

@@

For the linear density case, we can observe that the calculation of the 
potential and the $g_z$ component of its gradient needs a value of 
$D=1$ and $D=2$, respectively, in order to ensure an accuracy of 0.1\%.
On the other hand, the $g_{zz}$ computations show a noticeable 
difference between the thin and the thick shell: the first one needs a 
value of $D$ equal to 8 while the later only a $D$ of 3.5.
We will keep the more conservative result of $D$ equal to 8 in case of 
computing the Marussi tensor components.

@@

Lets consider a decreasing exponential density function that assumes 
the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's outer and 
inner radii, respectively, that can be defined as follows:

@@

Then, we proposed a thick spherical shell with a thickness of 35km and 
modelled by a mesh of $30^\circ \times 30^\circ$ tesseroids.
We have assigned the density function proposed on equation 
\ref{eq:density-exp} to our spherical shell model with a $b$ value of 
1km, i.e. a highly varying density function.

We computed the gravitational potential ($V$), the vertical component 
of its gradient ($g_z$) and the $g_zz$ component of the Marussi tensor 
on the big grid (see Table \ref{tab:grids}) by exploring different 
values of $D$ and $\delta$.
These results were used to obtain the percentage difference between 
them and their corresponding analytical solutions.

On Figure \ref{fig:grid-search} we show 
the resulting percentage difference for each gravitational field.
The points inside the dotted lines are the ones that present a 
numerical error lower than the 0.1\% threshold, i.e. they represent the 
eligible ($D$, $\delta$) pairs.
The one that produces the more efficient computation must be chosen as 
the default $D$ and $\delta$ value.

@@

Every point in the subregion with $\delta \le \delta_p$ (region B on 
Figure \ref{fig:D-delta-grid-search}) will have a numerical error lower 
than the 0.1\% threshold, but they will always produce more 
discretizations than the ($D_\text{linear}$, $\delta_p$) point leading 
to non-efficient computations.

@@

We can arrive to the same conclusion by analysing the computation 
efficiency: even if we find a point with acceptable numerical error in 
this region, we know it will produce more discretizations than the 
($D_\text{linear}$, $\delta_p$) point, leading to a non-efficient 
computation.
This is due to the fact that even an increase in $\delta$ will produce 
less density-based discretizations, the increase in $D$ will introduce 
even more subdivisions not only in the radial direction, but 
also in the latitude and longitude ones.

@@

We have considered the same spherical shell models we have used in the 
linear density case: a thin one (1km of thickness) and a thick one 
(35km of thickness) composed by $30^\circ \times 30^\circ$ tesseroids.
But in this case, we assigned an exponential density function like the 
one shown in equation \ref{eq:density-exp} to each shell, choosing 
different values of $b$ in order to explore distinct density 
variations.

Finally, we have computed the gravitational potential ($V$), the 
vertical component of its gradient ($g_z$) and the $g_zz$ component of 
the Marussi tensor on the big grid (see Table \ref{tab:grids}) and 
calculated the percentage difference between these numerical results and 
their corresponding analytical solutions.
Our intention is to set the default $\delta$ as the highest value whose 
numerical error is bellow the $0.1\%$ threshold.

>>>>

On the variable density tesseroid problem we found two main error 
sources in the GLQ that should be addressed: (a) the distance between 
the tesseroid and the computation point, and (b) 
the variation of the density function.

@@

\noindent Because the algorithm will be iteratively applied for every 
subdividing tesseroid, we'd like to note that this normalised density 
function will be the same for every subdividing tesseroid.
In case that the density function is constant, both the maximum and 
minimum densities are equal and the normalised function could not 
be defined, so the density-based discretization algorithm won't be 
applied.

@@

\noindent where the delta ratio $\delta$ is a predefined constant value 
and $L_r^\text{sm}/L_r$ is defined as the tesseroid size ratio.
Besides, $L_r^\text{sm}$ and $L_r$ are the radial dimensions of the 
tesseroid that is currently analysed and of the \emph{original} one, 
respectively.
On the first iteration, the former tesseroid corresponds to the 
\emph{original} one, so $L_r^\text{sm}/L_r$ is equal to one, but for 
future iterations this radial dimension will be computed as the 
difference of the top and bottom radii of the subdividing tesseroid.
The tesseroid size ratio is intended to reduce the number of probable 
unneeded discretizations and focus on discretizations that will 
significantly reduce the numerical error.
A big subdividing tesseroid with small maximum absolute difference would 
contribute more numerical error than a small subdividing tesseroid with 
higher maximum absolute difference.

@@

On the homogeneous density tesseroids case, \citet{Uieda2016} compared 
their numerical model with the analytical solution of a spherical 
shell in order to obtain default values for the distance-size ratio 
$D$.
We could follow this idea, but for our needs the spherical shell must 
have the same density function as the numerical model.
In contrast with the homogeneous density case, the analytical solutions 
for a variable density spherical shell are not present in the 
literature, so they must be obtained.

@@

The equation \ref{eq:shell-pot} allows to easily obtain the exact 
gravity potential generated by a spherical shell with a variable 
density in depth on any outside point located on the $z$ axe.
Due to the rotational symmetry along any axe that passes through the 
centre of the shell, the expression in equation \ref{eq:shell-pot} 
reproduces the gravity potential field on any outside point at distance 
$r$ from its centre.

@@

Firstly we need to solve the integral on equation \ref{eq:shell-pot} in 
order to obtain the analytical solution in this case.
Thus, the gravitational potential generated by a spherical shell with this 
density on any external point at distance $r$ from its centre can be 
obtained from the following expression:
\todo[inline]{revisar redaccion}

@@

\noindent The first term on equation \ref{eq:shell-pot-linear} 
reproduces the potential generated by a spherical shell with variable 
density $\rho(r') = ar'$, while the second one constitutes the 
potential generated by a spherical shell with homogeneous density $\rho 
= b$ \citep{Mikuska2006,Grombein2013}.

Secondly we can analyse how the density-based discretization algorithm 
works in case of a linear density function.
Because the absolute difference defined on equation 
\ref{eq:density-abs-diff} will always be zero, the discretization 
inequation \ref{eq:delta-density} will never be satisfied.
Therefore, independently of the linear function proposed, the 
density-based algorithm will never be applied, leaving the modified 
adaptive discretization algorithm as the only mechanism to control the 
accuracy of the numerical approximation.
That's why, for linear density scenarios, we only need to determine 
the minimum value of $D$ needed in order to guarantee an acceptable 
accuracy while ignoring the value of $\delta$.

@@

We will compute these gravitational fields on every point of the four 
grids for different values of $D$, ranging from 0.5 to 10 with a step 
of 0.5.
Then we will calculate the maximum difference between these results and the 
value that the analytical solution assumes at the same height of the 
computation grid.
Because the gravity fields generated by a tesseroid are approximated by 
the effect of point masses, the numerical results may vary between 
computation points at same height but at different latitude or 
longitude.
That's why we will choose for comparison the maximum difference between the 
results on every point of the grid and the analytical solution.

@@

For the linear density case, we can observe that the calculation of the 
potential and the $g_z$ component of its gradient needs a value of 
$D=1$ and $D=2$, respectively, in order to ensure a numerical error 
bellow the 0.1\% threshold.
On the other hand, the $g_{zz}$ computations show a noticeable 
difference between the thin and the thick shell: the first one needs a 
value of $D$ equal to 8 while the later only a $D$ of 3.5.
We will keep the more conservative result of $D$ equal to 8 in case of 
computing the Marussi tensor components.

@@

Firstly, lets consider a decreasing exponential density function that 
assumes the values of 2670 kg/m$^3$ and 3300 kg/m$^3$ on the shell's 
outer and inner radii, respectively, that can be defined as follows:

@@

Secondly, we propose a thick spherical shell with a thickness of 35km and 
modelled by a mesh of $30^\circ \times 30^\circ$ tesseroids in order to 
perform the exploration of the numerical error in the $D$, $\delta$ 
space.
We have assigned the density function proposed on equation 
\ref{eq:density-exp} to our spherical shell model with a $b$ value of 
1km, i.e. a highly varying density function.

Finally, we have computed the gravitational potential ($V$), the vertical 
component of its gradient ($g_z$) and the $g_{zz}$ component of the 
Marussi tensor on the big grid (see Table \ref{tab:grids}) by exploring 
different values of $D$ and $\delta$.
These results were used to obtain the percentage difference between 
them and their corresponding analytical solutions.

On Figure \ref{fig:grid-search} we show 
the resulting percentage difference for each gravitational field.
The points inside the dotted lines are the ones that present a 
numerical error lower than the 0.1\% threshold.
The one that produces the more efficient computation must be chosen as 
the default $D$ and $\delta$ value.

@@

Every point in the subregion with $\delta \le \delta_p$ (region B on 
Figure \ref{fig:D-delta-grid-search}) will have a numerical error lower 
than the 0.1\% threshold, but they will always produce more 
discretizations than the ($D_\text{linear}$, $\delta_p$) point leading 
to non-efficient computations.
It's worth remembering that a lower $\delta$ or a greter $D$ will 
produce more discretizations, increasing the accuracy but also the 
computation time.

@@

@@

Firstly, we have considered the same spherical shell models we have 
used in the linear density case: a thin one (1km of thickness) and a 
thick one (35km of thickness) composed by $30^\circ \times 30^\circ$ 
tesseroids.
But in this case, we assigned an exponential density function like the 
one shown in equation \ref{eq:density-exp} to each shell, choosing 
a few values of $b$ in order to explore different density 
variations.

Finally, we have computed the gravitational potential ($V$), the 
vertical component of its gradient ($g_z$) and the $g_{zz}$ component of 
the Marussi tensor on the big grid (see Table \ref{tab:grids}) and 
calculated the percentage difference between these numerical results and 
their corresponding analytical solutions.
Our intention is to set the default $\delta$ as the highest value whose 
numerical error is bellow the $0.1\%$ threshold.
